What We Eval in the Shadows

### 2021-04-17 paper/main.tex
varia

\begin{wraptable}{r}{7cm}\small\centering
  \small
  \centering
  \begin{tabular}{l|r|r|r|r}\hline
    Environments & \#sites & \%sites & \#funs. & \%funs. \\%
    \expandableinput tag/table-se-target-envs.tex
  \end{tabular}
  \caption{Target environments for \eval side-effects} \label{tab:se-env}
\end{wraptable}

Table~\ref{tab:se-env} shows the class of the environment where \eval side
effects happen (\emph{target environment}). It follows the classification
described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
represents a side-effect that happens in S4 or R6 object environments, and
\emph{Local} that corresponds to a side effect in local variables of the \eval
caller (offset 0 in Section~\ref{sec:env}). The table shows data for both \eval
sites and functions. For a function to have a given target environment class,
all of its \eval sites that do side effects must have the same target
environment class The majority of the \eval sites (\SESitesInOneClass) do all
side effects consistently in one environment class. The same happens at the
function level. Almost half of the sites (over a third of the functions) do side
effects in either \emph{Local} or \emph{Object} environments. This gives a ray
of hope for the compiler. Even though it is possible to do anything anywhere,
the data suggest that most side-effects are sane.

Table~\ref{tab:se-types} shows the proportion of the different side effects
that were recorded. In terms of calls, we see assignments primarily and in
terms of site definitions. This is expected. A subsequent \eval call will turn
a definition into an update. The most dangerous side effect is variable removal,
as it means that after a call to \eval some binding in some environments will
disappear. While rare, this happens, but the vast majority comes from a single
site called by the already mentioned \c{withr::execute\_handlers} function.
This makes sense as it is used to defer evaluation of an expression to after
the function exit and thus used for clean up. In the corpus, it is used almost
exclusively by the \c{tidyselect} package for removing the reference to the
current quosure environment while interpreting a data frame column
selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are the tools given to developers to
this end.
%This API is slightly more complex than just passing a string, it is
%conceivable that this may discourage some casual users. \Eval is also being
%used to reduce boilerplate code and provide convenience features for
%programmers.
In this section, we focus on the \emph{why}, giving real-world examples of the
\eval usage in the wild. All the examples are based on code from our corpus,
often simplified to fit the space and increase clarity. For each example, we
indicate the package and the function where it is located.

\subsection{Shapes of \eval}

% We start the discussion with the examples for the minimized expressions from
% Section~\ref{sec:minimized}.
In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
expressions passed to \eval. Here, we illustrate them on concrete instances.

@@

\subsection{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value. For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}. They support arithmetic operations and functions on real numbers such as \c{sin}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}
% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd

@@

% \subsubsection{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
% \c{match.call} walks up the call stack, captures the code that invoked the
% currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments. The pattern is to transform a call to some function \c{f} into a call to \c{g} with some arguments retained and others modified. This pattern is recognizable
% by the fact that the expression is a call and the target environment is that of
% the parent.

% % We have quantitative data on that if needed...
% This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model, \c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}. The R documentation of \c{match.call} explicitly describes this usage.

% In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame, to ensure that the rewritten call
% is evaluated in the same environment the original call was.

% \begin{lstlisting}
% Call <- match.call()
% tform <- Call[c(1,indx)]  # only keep the arguments we wanted
% tform[[1L]] <- quote(stats::model.frame)  # change the function called
% mf <- eval(tform, parent.frame())
% \end{lstlisting}

% A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This could also be achieved with \c{substitute}, but \c{match.call} does it in bulk.
% In package \emph{MedDietCalc},  in function \c{computeMDS95}, \c{match.call} retrieves the argument expressions of the call, which are evaluated in a dataframe \c{data} which contains the numeric data. Each argument of function \c{computeMDS95} is also a column in dataframe \c{data}. The advantage of using \c{eval} here is to be able to express an argument as a combination of several columns in the dataframe.
% \begin{lstlisting}
% computeMDS95 <- function (data, Vegetables, ...) {
% 	arguments <- as.list( match.call())
% 	Vegetables <- eval(arguments$Vegetables, data)
% \end{lstlisting}

% \c{substitute} can be used to capture only a part of the call, \ie one of the arguments. For instance, in package \emph{xyplot}, the expression for argument \c{subset} is captured with \c{substitute} and then evaluated in a dataframe \c{data} and in the definition environment of \c{x}.
% \begin{lstlisting}
% xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
% subset <- eval(substitute(subset), data, environment(x))
% \end{lstlisting}

% In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}. This pattern in relation with \c{model.frame} is also described explicitly in the R documentation of \c{match.call}.

% \subsubsection{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.

%  A call to \c{library(packageName)}, which loads libraries in R, for instance, will
% populate environment \c{package:packageName} using \eval. It is a \emph{base} function
% though, so we do not see it in the CRAN dataset. However, R still makes it
% possible to access any package environments, which we see in
% \packageNbPackageEnvPackages different packages in our corpus. Those packages perform
% introspection, to represent and search any environment by name, in package
% \emph{envnames}, to containerize tests, as with \emph{testthat} and
% \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
% implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.

% \subsubsection{Code Generation.} The more traditional use of \eval is to execute code
% that was assembled by the programmer into a string. Here we show the method
% \c{plot} for class \c{gmap} in package \c{gpmaps}, simplified for
% explanatory purposes.   \c{paste} is used to dispatch to a specialized plot function, among \c{plot2\_orig},  \c{plot3\_orig}, \c{plot3\_dec} and so on.
% \begin{lstlisting}
% eval(parse(text=paste('plot',gpmap$nloci,'_orig(gpmap)',sep="")))
% \end{lstlisting}

% In Kaggle, plots are often built with \eval, \c{paste} and \c{ggplot}. In this kernel for the \emph{digit-recognizer} challenge, \c{paste0} is used to generate \c{ggplot} commands featuring unique y labels to build 9 different graphs.

% \begin{lstlisting}
% p <- lapply(paste0('PC', 1:9), function(pc){
% 	eval(parse(text=paste0('ggplot(ld, aes(x=label, y=',pc,')) + geom_boxplot()')))
% })
% \end{lstlisting}

% The \parse and \eval combination is
% used by the \source function in the base library to load R code from a file
% in the current workspace.

%  Code generation is often used as a means to reduce boilerplate code;
% simple and repetitive code can easily be replaced with judicious use of \eval.
% For example, the \c{data.table} package uses \eval to calls the \c{options}
% function with named arguments taken from a vector of strings. While the benefits
% are limited in this example, it is an attractive tool for programmers.
% \begin{lstlisting}
%   opts = c("datatable.verbose"="FALSE", # ...many others
%   for (i in names(opts))
%     eval(parse(text=paste0("options(",i,"=",opts[i],")")))
% \end{lstlisting}
% This pattern is a special case of code generation, recognizable by the
% fact that \eval is executed in a loop.

% Using \c{parse}, \ie converting strings to R expressions, opens to a more flexible code generation, as the building blocks which are going to be concatenated together, do not need to be valid R expressions. On the contrary, combining expression created with \c{substitute} or \c{quote} requires more discipline. In Kaggle, which features less experienced programmers,  code generation for plots is widespread, and uses \c{parse} on strings with \eval.

% \todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp. But at the same type, we lack here an example of code generation using quite, enquote or substitute.}

% \todo{Add a pattern about Library Loading? \c{library(packageName)} in R uses \eval This is one of the patterns in Javascript...}

@@

\Eval in R is mostly used for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as
function or method call, or assignment to a local variable or an object
property. Compared to Javascript, \eval patterns in R we observed are much more
legitimate, although this is biased by the fact that our main corpus CRAN
features more experienced developers than  websites found at large on the web.
\todo{can we really claim that? SHould we talk about Kaggle there?}

\eval is a powerful tool and can be sometimes replaced with functions with more
specialized goals. For instance, variable lookup in any environment can be
performed with \c{get}, and assignment in any environment, with \c{assign}.
Building a call and executing can be done with \c{do.call}. Nevertheless, \eval
shines when expressions passed to it can be of many different types. % And it
is shorter to type, and no need to remember all the specialized functions, so
it is convenient.

@@

The origins of the expression passed to \eval, either constructed from language expressions, or stemming from reflection, or built by parsing strings. Reflection and constructed expressions dominate, which suggests the primary   use of \eval is meta-programming. % macros...

% how much can we say about the threat model?
Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them with other features that are more amenable to static
analysis. The data, unfortunately, suggests that replacing \eval is going to be
a lengthy journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated ways than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls
\eval with an expression parsed from string, can be categorized and often
replaced by equivalent yet more disciplined and safer code, the case of \eval
coming from reflection is much harder.

>>>>

\begin{wraptable}{r}{7cm}\small\centering
  \begin{tabular}{l|r|r|r|r}\hline
    Environments & \#sites & \%sites & \#funs. & \%funs. \\%
    \expandableinput tag/table-se-target-envs.tex
  \end{tabular}
  \caption{Target environments for side-effects} \label{tab:se-env}
\end{wraptable}

Table~\ref{tab:se-env} shows environment kinds where side-effects happen.
Function environments distinguish between \emph{local}, the caller environment
(offset 0), and \emph{function} (offset >0). \emph{Synthetic} represents
constructed environments. \emph{Object} is used to denote the environments
attached to objects and classes in the S4 and R6 object systems. \emph{Multiple}
denotes cases where side-effects to more than one kind originate from one site.
The table gives the number of call sites of \eval (and the ratio) performing
side-effect in a particular environment kind. The table also gives the number of
functions in which these sites occur (and ratio).

Most sites, \SESitesInOneClass, do all their side-effects consistently in one
kind of environments. The same happens at the function level. Almost half of the
sites (over a third of the functions) do side effects in either \emph{Local} or
\emph{Object} environments. This gives a ray of hope for a hypothetical R
compiler. Even though \eval can do anything anywhere, the data suggests most
effects are sane.

Table~\ref{tab:se-types} shows recorded effects. There are over 5M updates. In
terms of calls, we see assignments primarily and in terms of site definitions.
This is expected. A subsequent \eval call will turn a definition into an update.
The most dangerous side effect is variable removal, as it means that after a
call to \eval some binding in some environments will disappear. While rare, this
happens, but the vast majority comes from a single site called by the already
mentioned \c{withr::execute\_handlers} function. This makes sense as it is used
to defer evaluation of an expression to after the function exit and thus used
for clean up. In the corpus, it is used almost exclusively by the \c{tidyselect}
package for removing the reference to the current quosure environment while
interpreting a data frame column selectors.\footnote{\cf
\url{https://tidyselect.r-lib.org/}}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \c{substitute} and \c{eval} are the tools given to developers to
this end. In this section, we focus on the \emph{why}, giving real-world
examples of the \eval usage in the wild. All the examples are based on code from
our corpus, often simplified to fit the space and increase clarity. For each
example, we indicate the package and the function where it is located.

\subsection{Shapes redux}

In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
expressions passed to \eval. Here, we illustrate them on concrete instances.

@@

\subsection{Domain Specific Language.} R expressions are built and transformed
symbolically and then evalled to a concrete value. For instance, base R packages
\emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D}
and \c{deriv}. They support arithmetic operations and functions on real numbers
such as \c{sin}. In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear
Mixed Models), a \c{Dtensor} object is created, which contains expressions
derived twice (with \c{DD}) and then typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}

@@

@@

\Eval in R is mostly used for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as
function or method call, or assignment to a local variable or an object
property. Compared to Javascript, \eval patterns in R we observed are much more
legitimate, although this is biased by the fact that our main corpus CRAN
features more experienced developers than websites found at large on the web.

\eval is a powerful tool and can be sometimes replaced with functions with more
specialized goals. For instance, variable lookup in any environment can be
performed with \c{get}, and assignment in any environment, with \c{assign}.
Building a call and executing can be done with \c{do.call}. Nevertheless, \eval
shines when expressions passed to it can be of many different types.

@@

The origins of the expression passed to \eval, either constructed from language
expressions, or stemming from reflection, or built by parsing strings.
Reflection and constructed expressions dominate, which suggests the primary use
of \eval is meta-programming.

Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them with other features that are more amenable to static
analysis. The data, unfortunately, suggests that replacing \eval is going to be
a lengthy journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated ways than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls \eval
with an expression parsed from string, can be categorized and often replaced by
equivalent yet more disciplined and safer code, the case of \eval coming from
reflection is much harder.
